---
title: "Ovarian Cancer Risk Prediction Using `miRNA` Data"
format:
  revealjs: 
    scrollable: true
editor: visual
css: style.css

execute: 
  echo: false
---

## Objective

-- To predict ovarian cancer risk using miRNA expression profiles.

## Work done so far...

-- Preprocessed the RAW and Annotated datasets\
-- Implented TPM Normalization\
-- Built an initial LASSO linear regression model using glmnet and performed LOOCV\
-- Extracted the most informative features (using LASSO)\
-- Rebuilt a linear model with these selected features\
-- Extended this approach to logistic regression, decision trees, and PCA for comparison

```{r}

if (!require("pacman")) 
  install.packages("pacman")


pacman::p_load(sva,
               glmnet,
               caret,
               rpart.plot,
               randomForest,
               ggbiplot,
               tidyverse,
               reshape2)

```

```{r }
#Load in file with necessary functions/packages
source("../miRNAFunctions.R")
```

## What does the data look like? {.hscroll .scrollable}

(Post data-preprocessing and TPM normalization)

```{r}
data = readRDS("../../res/b4_oc_combat.rds")
head(data,7)
```

## Lasso 

::: panel-tabset

### Plot
```{r }
#Load in the labels for the 1st dataset
oc_ann = readRDS("../../data/blood_oc_full_ann.rds")
#Get the ovarian cancer sample IDs
oc_samples = oc_ann  %>% 
    filter(batch=="PMP" & (Histology=="Serous" & (STAGE=="Stage I/II" | STAGE=="Stage III/IV"))) %>%
    pull(Run)
#Get the healthy control sample IDs
control_samples = oc_ann %>% filter(batch=="NECC") %>% pull(Run)

#Load in the combined labels for the 2 datasets
ann_df = readRDS("../../data/b1b4OC_ann_full.rds") %>%
    mutate(score=case_when(sample_id %in% control_samples | Condition=="LR" ~0, #Label healthy control samples for model
                           sample_id %in% oc_samples ~1),   #Label cancer samples for model
          brca_status=case_when(Condition=="Control" | brca_status=="LR" ~"Healthy control",  #Label the samples for plotting
                                STAGE=="Stage I/II" ~"Stage I/II serous",
                                STAGE=="Stage III/IV" ~"Stage III/IV serous",
                                brca_status=="FH" ~"FH.b4", 
                                brca_status=="BRCA" ~"BRCA.b4",
                                brca_status=="Other" ~"Other.b4")) %>%
    mutate(brca_status = factor(brca_status, ordered=T, 
                                levels=c("Healthy control", "LR.b4", 
                                          "FH.b4", "BRCA.b4","Other.b4",
                                          "Stage I/II serous", "Stage III/IV serous")))  #reorder the plotting labels
     
```

```{r }
#Load in the data
df = readRDS("../../res/b4_oc_combat.rds") 
#Transpose table because glmnet package expects rows as samples, columns as features
df = df %>% t() %>% as.data.frame()

#Get the IDs of samples used to build model
train_samples = ann_df %>% filter(!is.na(score)) %>% pull(sample_id)

#Get table of samples used to build model
train_df = df[(rownames(df) %in% train_samples),]
#Get sample labels
y = ann_df %>% slice(match(rownames(train_df), sample_id)) %>%
    pull(score)
#Add column contianing the labels to table
train_df$y = y

#Build 1st model (LASSO)
mod = build_model(xy=train_df, alpha=1, 
                  nfolds = nrow(train_df), family = "gaussian", 
                  type.measure = "mse")
#Get the LOOCV error of the model
err = min(mod$cvm)
#Get the features selected by the model
nzc = extract_nzc(mod, lm=mod$lambda.min, family="gaussian")

#Subset the data to only the selected features
train_df = train_df %>% select(one_of(c(nzc, "y")))
#Build 2nd model using the selected features
mod2 <- lm(y ~ ., data = train_df)
#Get the adjusted r-squared of the 2nd model (shows goodness of fit)
r2 <- summary(mod2)$r.square
r2adj = summary(mod2)$adj.r.squared

#Subset the entire table to only the selected features
test_df = df %>% select(one_of(nzc))
#Use the 2nd model to predict on all samples
pred = predict(
            mod2,
            test_df) %>%
            as.data.frame() %>%
    rownames_to_column(var="sample_id") %>%
    dplyr::rename("pred"=".")
title = paste("OC risk predictions(Lasso)",
              "\nLOOCV error =", err,"\nR2 =",r2, "\nR2adj =", r2adj)

#Make boxplot
p = make_boxplot(pred, ann_df, x = "brca_status", y = "pred", 
                 title=title) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   #Rotate labels for spacing
print(p)

## To seeing the data in plotting the boxplot behind the scene
box_plot_data <- inner_join(pred, ann_df, by = join_by(sample_id))
```

### Selected Features
```{r}
nzc
```
:::



## Logistic Regression

::: panel-tabset

```{r}
#Get table of samples used to build model
train_df2 = df[(rownames(df) %in% train_samples),]
#Get sample labels
y2 = ann_df %>% slice(match(rownames(train_df2), sample_id)) %>%
    pull(score)

#Add column contianing the labels to table
train_df2$y2 = as.factor(y2)
test_df2 = df

# Convert predictors (features) to a matrix
X2 <- as.matrix(train_df2[, -which(names(train_df2) == "y2")])
```

### $\alpha$ = 0.5
```{r}
###### Cross validation on range of lambda \in (0,1) of length 100, while setting alpha = 0.5
set.seed(124)
t.cv <- 


logit_modelcv_lambda <- cv.glmnet(X2, y2, family = "binomial", type.measure = "class", alpha=0.5)
plot(logit_modelcv_lambda)
cat("min.cv.error:", min(logit_modelcv_lambda$cvm), "\nlambda of min.error:", logit_modelcv_lambda$lambda.min)
```

### $\lambda$ (default)
```{r}
####### Cross validation on range of alpha \in (0,1) of length 20, while setting lambda to default
logistic10cv <- function(data, alphas, num_folds = 10, seed = 124) {
  set.seed(seed)
  data <- data[sample(nrow(data)), ]  # Shuffle the data
  
  # Split data into 10 folds
  fold_size <- floor(nrow(data) / num_folds)
  folds <- split(data, rep(1:num_folds, each = fold_size, length.out = nrow(data)))
  
  # Prepare to store CV errors
  cv_errors <- matrix(0, nrow = length(alphas), ncol = num_folds)
  
  for (i in 1:num_folds) {
    # Create training and validation sets
    validation_data <- folds[[i]]
    training_data <- do.call(rbind, folds[-i])
    
    # Separate features and labels
    X_train <- as.matrix(training_data[, -ncol(training_data)])
    y_train <- as.factor(training_data$y)
    X_val <- as.matrix(validation_data[, -ncol(validation_data)])
    y_val <- as.factor(validation_data$y)
    
    for (j in 1:length(alphas)) {
      alpha_val <- alphas[j]
      
      # Fit logistic regression with specified alpha
      logit_model <- glmnet(X_train, y_train, family = binomial, alpha = alpha_val)
      
      # Predict probabilities on validation set
      val_predictions <- predict(logit_model, newx = X_val, type = "response")
      
      # Convert probabilities to class labels (threshold 0.5)
      pred_class <- ifelse(val_predictions > 0.5, 1, 0)
      pred_class <- as.vector(pred_class[, ncol(pred_class)])  # Select the last column
      
      # Calculate error rate for the current fold
      cv_errors[j, i] <- mean(pred_class != as.numeric(y_val) - 1)
    }
  }
  
  # Calculate mean CV errors and standard errors for each alpha
  mean_cv_errors <- rowMeans(cv_errors)
  std_errors <- apply(cv_errors, 1, sd) / sqrt(num_folds)
  
  return(list(alphas = alphas, mean_cv_errors = mean_cv_errors, std_errors = std_errors))
}

# Define a range of alpha values to test
alphas <- seq(0, 1, by = 0.1)

# Perform 10-fold cross-validation for different alpha values
cv_results <- logistic10cv(train_df2, alphas)

# Plot the 10-fold CV error curve with standard error bars
plot(cv_results$alphas, cv_results$mean_cv_errors, type = "b", pch = 19, xlab = "alpha", ylab = "10-Fold CV Error",
     main = "10-Fold CV Error Curve for Logistic Regression")
segments(cv_results$alphas, cv_results$mean_cv_errors - cv_results$std_errors,
         cv_results$alphas, cv_results$mean_cv_errors + cv_results$std_errors, col = "blue")
legend("topright", legend = "CV Error +/- 1 SE", col = "blue", lty = 1, pch = 19)

data.frame(cv_results)
```

### Box-plot
```{r}
# Normal Logistic regression model
logit_model2 <- glmnet(X2, y2, 
                   family = binomial, lambda =0.004,  alpha = 1)


# Predict on test data
predictions2 <- predict(logit_model2, newx = as.matrix(test_df2), type = "response")
predictions2_train_df2 <- predict(logit_model2, newx = X2, type = "response")

# Convert probabilities to class labels (threshold 0.5)
pred_class2 <- ifelse(predictions2 > 0.5, 1, 0)
pred_class2 <- pred_class2[, ncol(pred_class2)]
pred_class_train_df2 <- ifelse(predictions2_train_df2 > 0.5, 1, 0)
pred_class_train_df2 <- as.vector(pred_class_train_df2[, ncol(pred_class_train_df2)])

# Evaluate the model
#confusionMatrix(factor(pred_class), factor(test_df$Species))
confusionMatrix(factor(pred_class_train_df2), train_df2$y2)


pred2 = as.data.frame(predictions2[,ncol(predictions2)]) %>%
    rownames_to_column(var="sample_id") %>% dplyr::rename(pred = "predictions2[, ncol(predictions2)]")
title = paste("OC risk predictions(Logistic)",
              "\nAccuracy =",100, "\nerror =", logit_model2$jerr)

#Make boxplot
p = make_boxplot(pred2, ann_df, x = "brca_status", y = "pred", 
                 title=title) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   #Rotate labels for spacing
print(p)

# plot data
box2_plot_data <- inner_join(pred2, ann_df, by = join_by(sample_id))
```

:::



## Decision Tree

::: panel-tabset

### DT

```{r}
#Get table of samples used to build model
train_df3 = df[(rownames(df) %in% train_samples),]
#Get sample labels
y3 = ann_df %>% slice(match(rownames(train_df3), sample_id)) %>%
    pull(score)
#Add column contianing the labels to table
train_df3$y3 = as.factor(y3)
test_df3 = df

#Get table of samples used to build model
test_samples3 = ann_df %>% filter(is.na(score)) %>% pull(sample_id)
test_df3b = df[(rownames(df) %in% test_samples3),]


# Train the decision tree model
tree_model <- rpart(train_df3$y3~., 
                    data = train_df3[,1:ncol(train_df3)-1], 
                    method = "class")  # 'class' for classification

# Summary of the model (prints the decision tree)
print(tree_model)

# Visualize the decision tree
rpart.plot(tree_model, type = 2, extra = 102, fallen.leaves = TRUE)

# Predict on test data
predictions3 <- predict(tree_model, newdata = test_df3, type = "class")
predictions3b <- predict(tree_model, newdata = test_df3b, type = "class")
predictions_train_df3 <- predict(tree_model, newdata = train_df3, type = "class")

# Evaluate the model with a confusion matrix
confusionMatrix(predictions_train_df3, train_df3$y3)
```

```{r}
print(as.data.frame( tree_model$variable.importance))

# Plot the variable importance
barplot(tree_model$variable.importance,
        main = "Variable Importance from Decision Tree",
        xlab = "Features",
        ylab = "Importance",
        col = "steelblue",
        horiz = FALSE,
        las = 2)  # Rotate axis labels for better readability
```

### Box-plot

```{r}
#Get the LOOCV error of the model
err3 = min(tree_model$cptable[,3])
#Get the features selected by the model
nzc3 = names(tree_model$variable.importance)

#Subset the data to only the selected features
train_df3 = train_df3 %>% select(one_of(c(nzc3, "y3")))
train_df3$y3 <- as.numeric(as.character(train_df3$y3))
#Build 2nd model using the selected features
lm_model3 <- lm(y3 ~ ., data = train_df3)
#Get the adjusted r-squared of the 2nd model (shows goodness of fit)
r2 <- summary(lm_model3)$r.square
r2adj = summary(lm_model3)$adj.r.squared

#Subset the entire table to only the selected features
test_df3 = df %>% select(one_of(nzc3))
#Use the 2nd model to predict on all samples
pred3 = predict(
            lm_model3,
            test_df3) %>%
            as.data.frame() %>%
    rownames_to_column(var="sample_id") %>%
    dplyr::rename("pred"=".")
title = paste("OC risk predictions(DT)",
              "\nLOOCV error =", err3, "\nR2 =",r2, "\nR2adj =", r2adj)

#Make boxplot
p = make_boxplot(pred3, ann_df, x = "brca_status", y = "pred", 
                 title=title) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   #Rotate labels for spacing
print(p)

# plot data
box3_plot_data <- inner_join(pred3, ann_df, by = join_by(sample_id))
```
:::





## PCA


::: panel-tabset


### Train_df
```{r }
# library(ggbiplot)
#Get table of samples used to build model
train_df4 = df[(rownames(df) %in% train_samples),]
y4 = ann_df %>% slice(match(rownames(train_df4), sample_id)) %>%
    pull(score)  #Get sample labels
train_df4$y4 = as.factor(y4)    #Add column contianing the labels to table
test_df4 = df

# PCA
pc_train_df <- prcomp(train_df4[,1:ncol(train_df4)-1],
             center = TRUE,
             scale. = TRUE)
# print(pc_train_df)
summary(pc_train_df)
plot(pc_train_df, type = "lines")

# Bi-Plot
g <- ggbiplot(pc_train_df, 
              obs.scale = 1, 
              var.scale = 1, 
              groups = train_df4$y4, 
              ellipse = TRUE, 
              circle = TRUE,
              varname.abbrev = TRUE,
              var.factor = 2,
              varname.size = 0,
              ellipse.prob = 0.90)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)
```


### PC (1-10)
```{r}
# Prediction with Principal Components
pc_train4_pred <- predict(pc_train_df, train_df4)
pc_train4_pred <- data.frame(pc_train4_pred, y4=as.numeric(as.character(
  train_df4[,ncol(train_df4)])))
pc_test4_pred <- predict(pc_train_df, test_df4)
pc_test4_pred <- data.frame(pc_test4_pred)

# Select only the desired features and target
subset_train_df10 <- pc_train4_pred %>%
  select(y4, PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10)

subset_train_df35 <- pc_train4_pred %>%
  select(y4, PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10,
         PC11, PC12, PC13, PC14, PC15, PC16, PC17, PC18, PC19, PC20,
         PC21, PC22, PC23, PC24, PC25, PC26, PC27, PC28, PC29, PC30,
         PC31, PC32, PC33, PC34, PC35)

subset_test_df10 <- pc_test4_pred %>%
  select(PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10)

subset_test_df35 <- pc_test4_pred %>%
  select(PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10,
         PC11, PC12, PC13, PC14, PC15, PC16, PC17, PC18, PC19, PC20,
         PC21, PC22, PC23, PC24, PC25, PC26, PC27, PC28, PC29, PC30,
         PC31, PC32, PC33, PC34, PC35)


# Fit the model for the first 10 PC's
lm_model4 <- lm(y4 ~ ., data = subset_train_df10)
#Get the adjusted r-squared of the 2nd model (shows goodness of fit)
r2 <- summary(lm_model4)$r.square
r2adj = summary(lm_model4)$adj.r.squared


#Use the 2nd model to predict on all samples
pred4 = predict(
            lm_model4,
            subset_test_df10) %>%
            as.data.frame() %>%
    rownames_to_column(var="sample_id") %>%
    dplyr::rename("pred"=".")
title = paste("OC risk predictions(PC1-10)", "\n72% Var. Exp",
               "\nR2 =",r2, "\nR2adj =", r2adj)

#Make boxplot
p = make_boxplot(pred4, ann_df, x = "brca_status", y = "pred", 
                 title=title) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   #Rotate labels for spacing
print(p)
# plot data
box4a_plot_data <- inner_join(pred4, ann_df, by = join_by(sample_id))





# Fit the model for the first 35 PC's
lm_model4 <- lm(y4 ~ ., data = subset_train_df35)
#Get the adjusted r-squared of the 2nd model (shows goodness of fit)
r2 <- summary(lm_model4)$r.square
r2adj = summary(lm_model4)$adj.r.squared


#Use the 2nd model to predict on all samples
pred4 = predict(
            lm_model4,
            subset_test_df35) %>%
            as.data.frame() %>%
    rownames_to_column(var="sample_id") %>%
    dplyr::rename("pred"=".")
title = paste("OC risk predictions(PC1-35)", "\n94% Var. Exp",
               "\nR2 =",r2, "\nR2adj =", r2adj)



```

### PC (1-35)
```{r}
#Make boxplot
p = make_boxplot(pred4, ann_df, x = "brca_status", y = "pred", 
                 title=title) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   #Rotate labels for spacing
print(p)


```
:::


## PCA with Lasso selected variables



::: panel-tabset

### bi and scree

```{r }
library(ggbiplot)
#Get table of samples used to build model
train_df5 = train_df
test_df5 = df

# PCA
pc_train_df2 <- prcomp(train_df5[,1:ncol(train_df5)-1],
             center = TRUE,
             scale. = TRUE)
# print(pc_train_df2)
summary(pc_train_df2)
plot(pc_train_df2, type = "lines")

# Bi-Plot
g <- ggbiplot(pc_train_df2, 
              obs.scale = 1, 
              var.scale = 1, 
              groups = as.factor(train_df5$y), 
              ellipse = TRUE, 
              circle = TRUE,
              varname.abbrev = TRUE,
              var.factor = 2,
              varname.size = 3,
              ellipse.prob = 0.95)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)
```


## Box 
```{r}
# Prediction with Principal Components
pc_train4_pred <- predict(pc_train_df, train_df4)
pc_train4_pred <- data.frame(pc_train4_pred, y4=as.numeric(as.character(
  train_df4[,ncol(train_df4)])))
pc_test4_pred <- predict(pc_train_df, test_df4)
pc_test4_pred <- data.frame(pc_test4_pred)

# Select only the desired features and target
subset_train_df10 <- pc_train4_pred %>%
  select(y4, PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10)

subset_train_df35 <- pc_train4_pred %>%
  select(y4, PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10,
         PC11, PC12, PC13, PC14, PC15, PC16, PC17, PC18, PC19, PC20,
         PC21, PC22, PC23, PC24, PC25, PC26, PC27, PC28, PC29, PC30,
         PC31, PC32, PC33, PC34, PC35)

subset_test_df10 <- pc_test4_pred %>%
  select(PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10)

subset_test_df35 <- pc_test4_pred %>%
  select(PC1, PC2, PC3, PC4, PC5, PC6, PC7, PC8, PC9, PC10,
         PC11, PC12, PC13, PC14, PC15, PC16, PC17, PC18, PC19, PC20,
         PC21, PC22, PC23, PC24, PC25, PC26, PC27, PC28, PC29, PC30,
         PC31, PC32, PC33, PC34, PC35)


# Fit the model for the first 10 PC's
lm_model4 <- lm(y4 ~ ., data = subset_train_df10)
#Get the adjusted r-squared of the 2nd model (shows goodness of fit)
r2 <- summary(lm_model4)$r.square
r2adj = summary(lm_model4)$adj.r.squared


#Use the 2nd model to predict on all samples
pred4 = predict(
            lm_model4,
            subset_test_df10) %>%
            as.data.frame() %>%
    rownames_to_column(var="sample_id") %>%
    dplyr::rename("pred"=".")
title = paste("OC risk predictions(PC1-10)", "\n72% Var. Exp",
               "\nR2 =",r2, "\nR2adj =", r2adj)

#Make boxplot
p = make_boxplot(pred4, ann_df, x = "brca_status", y = "pred", 
                 title=title) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   #Rotate labels for spacing
print(p)
# plot data
box4a_plot_data <- inner_join(pred4, ann_df, by = join_by(sample_id))


# Fit the model for the first 35 PC's
lm_model4 <- lm(y4 ~ ., data = subset_train_df35)
#Get the adjusted r-squared of the 2nd model (shows goodness of fit)
r2 <- summary(lm_model4)$r.square
r2adj = summary(lm_model4)$adj.r.squared


#Use the 2nd model to predict on all samples
pred4 = predict(
            lm_model4,
            subset_test_df35) %>%
            as.data.frame() %>%
    rownames_to_column(var="sample_id") %>%
    dplyr::rename("pred"=".")
title = paste("OC risk predictions(PC1-35)", "\n94% Var. Exp",
               "\nR2 =",r2, "\nR2adj =", r2adj)

#Make boxplot
p = make_boxplot(pred4, ann_df, x = "brca_status", y = "pred", 
                 title=title) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))   #Rotate labels for spacing
print(p)

# plot data
box4b_plot_data <- inner_join(pred4, ann_df, by = join_by(sample_id))
```
:::



## Predictions
```{r}

# plot data
box4b_plot_data <- inner_join(pred4, ann_df, by = join_by(sample_id))

models_pred <- data.frame(box_plot_data[,1:2],
                                DT_pred=box3_plot_data[,2],
                                pc1_10_pred=box4a_plot_data[,2],
                                pc1_35_pred=box4b_plot_data[,2],
                                Logst_pred=box2_plot_data[,2],
                                box_plot_data[3:4])
models_pred <- models_pred %>%dplyr::rename(Lasso_pred = pred)
FH.b4_models_pred <- models_pred %>% filter(brca_status == "FH.b4")

# Line graph to comparison of predictions base on each models
matplot(t(FH.b4_models_pred[2:6]), type = "b", pch = 16, lty = 1, col = 1:10, xaxt = "n", xlab = "Columns", ylab = "predictions", 
        main = "Line Graph for FH.b4 model's preditions")
axis(1, at = 1:5, labels = colnames(FH.b4_models_pred)[2:6])
legend("topright", legend = FH.b4_models_pred$sample_id, col = 1:10, lty = 1, pch = 16, cex = 0.8)


# Line graph to comparison of predictions base on each models without legend
matplot(t(FH.b4_models_pred[2:6]), type = "b", pch = 16, lty = 1, col = 1:10, xaxt = "n", xlab = "Columns", ylab = "predictions", 
        main = "Line Graph for FH.b4 model's preditions")
axis(1, at = 1:5, labels = colnames(FH.b4_models_pred)[2:6])
```

```{r}
# Plot of CV error of each model
cv_error_plot_data <- c(min(logit_modelcv_lambda$cvm), err3,err)
x_labels <- c("Logistic", "DT", "Lasso")
plot(cv_error_plot_data,type = "b", pch = 19, xaxt = "n", ylab = "Y Values", xlab = "Custom X Axis", main = "CV Error of Each Model")

# Add custom x-axis with character labels
axis(1, at = 1:length(x_labels), labels = x_labels)
```


